Modelling BCG vaccination in the UK

### 2019-07-25 rmds/01-introduction.Rmd
added tb-epi changes from Hannah
tb-epi updates + intro link to formats
outlined model results

# Introduction

@@

This thesis explores the impact of changing BCG policy in England, with the aim of informing global efforts to control TB. As a first step I outline some of the key background information and discuss the tooling that I developed in order to explore this information more fully. I then make use of the detailed PHE routine TB surveillance data to explore the current epidemiology of TB in England. Next, I use statistical models that make use of this data to explore the impact from the 2005 change in BCG policy. Finally I develop, and fit, a detailed, semi-sthochastic, mechanistic model of TB and BCG vaccination in England in order to forecast the ongoing impact of the change in policy versus multiple alternative scenarios. The remainder of this chapter outlines: the theoretical framework used in this thesis; the aims and objects that were used to motivate this thesis; the chapter structure of this thesis; and the output from this thesis.

@@

* Abbott S, Christensen H, Lalor MK, Zenner D,Campbell C, Ramsay M, Brooks-Pollock E. *Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system*, doi: https://doi.org/10.1101/366476

@@

* **`idmodelr`**: An R package that contains a library  of infectious disease models as well as modelling utilities. It provides tooling that includes: example SEI/SEIR/SHLIR/SHLITR model code, a model solving wrapper; a model summary function; a scenario analysis function). Used by the explore infectious disease model dashboard (http://seabbs.co.uk/shiny/exploreidmodels/) for all functionality. Available from GitHub using `devtools::install_github("seabbs/idmodelr")`. Link: https://www.samabbott.co.uk/idmodelr/

>>>>

# Introduction {#introduction}

@@

This thesis explores the impact of changing BCG policy in England, with the aim of informing global efforts to control TB. As a first step I outline some of the key background information and discuss the tooling that I developed in order to explore this information more fully. I then make use of the detailed PHE routine TB surveillance data to explore the current epidemiology of TB in England. Next, I use statistical models that make use of this data to explore the impact from the 2005 change in BCG policy. Finally I develop, and fit, a detailed, semi-sthochastic, mechanistic model of TB and BCG vaccination in England in order to forecast the ongoing impact of the change in policy versus multiple alternative scenarios. This thesis is also available as a website^[Website: https://www.samabbott.co.uk/thesis/], a pdf^[PDF: https://www.samabbott.co.uk/thesis/thesis.pdf], and a reproducible Rmarkdown^[Rmarkdown: https://github.com/seabbs/thesis]

The remainder of this chapter outlines: the theoretical framework used in this thesis; the aims and objects that were used to motivate this thesis; the chapter structure of this thesis; and the output from this thesis.

@@

* Abbott, S., Christensen, H., Lalor, M. K., Zenner, D., Campbell, C., Ramsay, M. E., & Brooks-Pollock, E. (2019). *Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: Observational study using the Enhanced Tuberculosis Surveillance system*. Vaccine, 1–6. doi: http://doi.org/10.1016/j.vaccine.2019.06.056 (preprint: https://doi.org/10.1101/366476)

@@

* **`idmodelr`**: An R package that contains a library  of infectious disease models as well as modelling utilities. It provides tooling that includes: example SEI/SEIR/SHLIR/SHLITR model code, a model solving wrapper; a model summary function; a scenario analysis function). Used by the explore infectious disease model dashboard (http://seabbs.co.uk/shiny/exploreidmodels/) for all functionality. Available from CRAN using `install.packages("idmodelr")` or install the development version from GitHub with `devtools::install_github("seabbs/idmodelr")`. Link: https://www.samabbott.co.uk/idmodelr/

### 2019-07-25 rmds/08-model-development.Rmd
outlined model results

This chapter has outlined a realistic dynamic transmission model of TB that includes key features required to investigate BCG vaccination policy and is robustly parameterised from an extensive, and previously unused in a TB model, routine surveillance data-set. Transformations, and approximations, of parameters have been used to make the best use of available data. However, there are several key limitations. Firstly, the model presented here does not explicitly model TB transmission in the non-UK born. This means that in order to initialise the model assumptions must be made about the historic number of non-UK born cases and the future incidence in the non-UK born must also be assumed in order to produce projections of future TB incidence. However, this simplification allows many complexities of TB in the non-UK born to be discounted, such as the rate of case importation, heterogeneity amongst the non-UK born from different countries, and mixing within the non-UK born. Secondly, the model presented here does not include high and low risk stratification within the UK born. Individuals that are from countries with incidence above 40 per 100,000, or that have parents/grandparents from countries with incidence above 40 per 100,000, are considered at higher risk of TB.[@PublicHealthEngland2011a] In addition, individuals living in areas of the UK with incidence above this threshold are also considered at higher risk. Current BCG vaccination policy targets high risk neonates for vaccination, with low risk neonates not being vaccinated. Ideally, this high/low risk stratification would be included as it would allow the evaluation of the current BCG vaccination policy. This has not been possible as there is little data from which to extrapolate either the number of high risk notifications in the ETS, or the size of the high risk population. There is also little evidence to suggest the degree of mixing between the non-UK born, the high risk UK born population and the low risk non-UK born population. It is likely that introducing this structure into the model, without the data outlined above, would lead to the model being poorly specified and therefore failing to fit to the observed data. Instead, in the final chapter in this thesis, the high risk neonatal programme will be proxied by a universal neonatal programme. This will allow for a comparison to be made between school-age and neonatal vaccination but does not allow for the impact of targeting high risk individuals to be evaluated. Finally, the model presented here does not include the full complexity of TB epidemiology. Drug resistant TB may have more severe outcomes, standard TB treatment may fail resulting in a longer period on treatment, and TB outcomes may vary by risk group.[@PHE2017] However, drug resistant TB cases are known to make up a small fraction of TB cases in England in the UK born, and variable treatment times have been included in the prior distribution of treatment times and TB outcomes. Model parameters have also been stratified by pulmonary status and age group where appropriate. Additionally, complexity has been included for the action of the BCG vaccine, with realistic waning in effectiveness. Observed age-specific mortality rates and the number of live births has also been included, allowing for realistic population demographics. This means that estimates of the impact of the change in vaccination policy are likely to be more accurate, whereas a more complex model of the epidemiology of TB would likely have little impact on these results.

>>>>

This chapter has outlined a realistic dynamic transmission model of TB that includes key features required to investigate BCG vaccination policy and is robustly parameterised from an extensive, and previously unused in a TB model, routine surveillance data-set. Transformations, and approximations, of parameters have been used to make the best use of available data. However, there are several key limitations. Firstly, the model presented here does not explicitly model TB transmission in the non-UK born. This means that in order to initialise the model assumptions must be made about the historic number of non-UK born cases and the future incidence in the non-UK born must also be assumed in order to produce projections of future TB incidence. However, this simplification allows many complexities of TB in the non-UK born to be discounted, such as the rate of case importation, heterogeneity amongst the non-UK born from different countries, and mixing within the non-UK born. Secondly, the model presented here does not include high and low risk stratification within the UK born. Individuals that are from countries with incidence above 40 per 100,000, or that have parents/grandparents from countries with incidence above 40 per 100,000, are considered at higher risk of TB.[@PublicHealthEngland2011a] In addition, individuals living in areas of the UK with incidence above this threshold are also considered at higher risk. Current BCG vaccination policy targets high risk neonates for vaccination, with low risk neonates not being vaccinated. Ideally, this high/low risk stratification would be included as it would allow the evaluation of the current BCG vaccination policy. This has not been possible as there is little data from which to extrapolate either the number of high risk notifications in the ETS,[@PHE2018] or the size of the high risk population. There is also little evidence to suggest the degree of mixing between the non-UK born, the high risk UK born population and the low risk non-UK born population. It is likely that introducing this structure into the model, without the data outlined above, would lead to the model being poorly specified and therefore failing to fit to the observed data. Instead, in the final chapter in this thesis, the high risk neonatal programme will be proxied by a universal neonatal programme. This will allow for a comparison to be made between school-age and neonatal vaccination but does not allow for the impact of targeting high risk individuals to be evaluated. Finally, the model presented here does not include the full complexity of TB epidemiology. Drug resistant TB may have more severe outcomes, standard TB treatment may fail resulting in a longer period on treatment, and TB outcomes may vary by risk group.[@PHE2017] However, drug resistant TB cases are known to make up a small fraction of TB cases in England in the UK born, and variable treatment times have been included in the prior distribution of treatment times and TB outcomes. Model parameters have also been stratified by pulmonary status and age group where appropriate. Additionally, complexity has been included for the action of the BCG vaccine, with realistic waning in effectiveness. Observed age-specific mortality rates and the number of live births has also been included, allowing for realistic population demographics. This means that estimates of the impact of the change in vaccination policy are likely to be more accurate, whereas a more complex model of the epidemiology of TB would likely have little impact on these results.

### 2019-07-25 rmds/10-model-results.Rmd
checked accroynm usage
outlined model results

* Why
* What did I do in the last chapter and how does this extend it
* What I did summary

## Method

@@

* Outline scenarios considered

### Future Non-UK born cases assumption

* Exponential decay
* Remain constant
* Compare and constrast what these mean at different timelines

@@

## Impact between 2005 and 2010

  * Comparision of cases at multiple time horizons.
  * Comparision of age distribution of cases.
  * Comparision of deaths at multiple time horizons.
  * Comparision of age distribution of deaths.
  * All in reflection of reality.

## Non-UK born cases decline exponentially over time

## Non-UK born cases stay at todays levels

@@

* Summary of work from chapter
* Strengths and weaknesses
* Literature comparisions
* Concliusions

@@

* Best vaccination scenario according to model
* Implications for actual policy
*

>>>>

In the previous chapter I outlined a model fitting pipeline and discussed the results from using it on the model developed in Chapter \@ref(model-development). Whilst this fitted model may be used to explore the epidemiology of tuberculosis (TB) in the early 2000's it does not - as currently stands - explore the impact of the 2005 change in BCG vaccination policy (Chapter \@ref(background)).

This chapter details the approach used to extrapolate the findings from the previous chapter beyond the change in BCG policy in 2005 and into the future. It first outlines the scenarios considered, then details the assumptions used to expand the time horizon of the model. Finally the impact of each scenario is explored over multiple time horizons. As discussed in the previous chapter these findings are preliminary in nature, meaning quantative conclusions cannot be drawn and qualative conclusions must be appropriately caveated.

## Methods

@@

I considered three scenarios from 2005 onwards. These were:

* Universal BCG vaccination of those at school-age continued with the same coverage as previously.
* Universal BCG vaccination of those at school-age was phased out in 2005 and replaced with universal BCG vaccination of neonates with the same coverage levels as assumed for the BCG schools scheme.
* Universal BCG vaccination of those at school-age was phased out in 2005.

Universal vaccination of neonates was used as a proxy for targeted vaccination of high risk neonates. This was necessary because the high risk population was not modelled in the model developed in Chapter \@ref(model-development) due to the lack of data on which to base key assumptions. No vaccination was used as a baseline in order to explore the absolute impact of vaccination. Vaccination coverage was assumed to be constant across all scenarios as there was little data on which to base between assumption variation. Regardless of the scenario considered it was assumed that school-age vaccination was in place from 1953 through to 2004.

### Forecasting assumptions

Data on non-UK born cases, which were imported into the model via the force of infection (Chapter \@ref(model-development)), were not avialable beyond 2015. To account for this an, age and year adjusted, poisson regression model was used to forecast future TB incidence in the non-UK born with age treated as a categorical variable. As for years with data uncertainty was introduced into these forecasts by assuming that non-UK born incidence rates were scaled using the fitted measurement error and normally distributed with a standard distribution based on the fitted measurement standard error (Chapter \@ref(model-fitting)).

As outlined in Chapter \@ref(model-development), births from 2015 onwards were based on projections from the Office for National Statistics (ONS). Age-specific mortality rates were estimated for 2016 onwards using ONS estimates from 1981-2015, and an exponential model (Chapter \@ref(model-development). Both births and age-specific mortality rates were assumed to have a normal distribution with a standard deviation of 5% of the proect value. It was assumed that all other parameters were unchanged from the values estimated for 2000-2004 (Chapter \@ref(model-fitting).

### Analytical methods

The estimated age-stratified, and aggregated, TB incidence rates were compared both visually and numerically from 2005 through to 2040 for each scenario. From 2005-2015 observed data were also compared to the estimates from each scenario. Multiple time horizons were evaluated as intitially the impact of any policy change may be masked by the large resevoir of vaccinated individuals in the populations and because of the impact of the assumed decrease in non-UK born cases over time.

@@

The results presented in the following section should be considered preliminary because of the low quality of fit achieved in Chapter \@ref(model-fitting). This means that quantative estimates are unlikely to be accurate. However, the underlying changes in dynamics caused by vaccination may be used for insight into the likely impact of each scenario and therefore there is still value in exploring these results.

## Comparing vaccination scenarios with observed data

  * Comparision of cases.
  * Comparision of age distribution of deaths.
  * All in reflection of reality.

## Forecasting the long-term impact of each vaccination scenario.

@@

* Summary of work from chapter
* Strengths and weaknesses
* Literature comparisions
* Further work
* Conclusions

### Summary

The preliminary results presented in this chapter indicate that continuing BCG vaccination of those at school-age would have ...

They also indicate that universal neonatal vaccination would have ...

In comparison stopping all BCG vaccination would have resulted in ...

Over the longer term these results indicate that school-age BCG vaccination would have ... in comparison to neonatal vaccination. Both scenarios in *placeholder* in comparison to no vaccination.

### Strengths and weaknesses

The model developed in Chapter \@ref(model-development) was motivated by existing theory and robustly parameterised to the available data. It represents the only open-source model of TB transmission and BCG vaccination. However the model fitting pipeline developed in Chapter \@ref(model-fitting) did not produce a good fit to the observed data. Ad-hoc model calibration (as discussed in Chapter \@ref(model-fitting)) failed to significantly improve on this fit. This means that the findings presented in this chapter can be considered as indictative only. However, these findings still represent the only modelling study of TB dynamics after a large scale change in BCG vaccination policy. The lack of data to support modelling the high-risk TB population population meant that targeted vaccination of high-risk neonates could not be considered as a scenario. This means that the results presented in this chapter do not contain the vaccination policy that is currently in place and so findings from the model cannot be directly compared to observed incidence data. However, only considering scenarios that alter the age of those vaccinated, rather than both the age of those vaccinated and the targeting of the vaccine, make understanding the impact of changes in vaccination policy easier to determine. The forecasts presented in this chapter are highly sensitive to the forecasted number of non-UK born TB cases. Whilst the regression method outlined in this chapter extrapolates based on current age-stratified trends it may be the case that this extrapolation breaks down over the long - or short - term. To a lesser extent the forecasts presented here are sensitive to the projected number of births and mortality rates. This is particularly the case for incidence rates in neonates and in older adults.

### Literature + Thesis comparisons

### Further work

Further work is need to improve the fit of the model to observed data. This will result in improved forecasts. In addition additional vaccination coverage scenarios could be considered that explored the impact of vaccinating a reduced proportion of both those at school-age and neonates. It additional data becomes available, or with the appropriate assumptions, the inclusion of the high-risk population into the model would allow the evaluate of targeted high-risk neonatal vaccination in comparison to the other scenarios considered here. The extrapolation of the trend non-UK born cases is a limitation of this model and as such should be further explored using other assumptions such as constant non-UK born cases, incidence rates based on expert opinion and estimates based on other global modelling studies.

### Conclusions

@@

* Best vaccination scenario according to model
* Implications for actual policy
